\chapter{Setting and Varying Flags}
\label{flags}

%Some facilities which can be helpful in finding modes
%appropriate for proving particular theorems automatically are
%discussed in section \ref{modes}.  





The settings of flags are very important during automatic search.  For
example, flags like \indexflag{REWRITE-DEFNS} and
\indexflag{REWRITE-EQUALITIES} must be set to T (true) if you want
definitions and equalities, respectively, to be rewritten when the
expansion tree is created from the formula to be proved.  To see all
of the flags which will affect mating-search, check the facilities
guide.

Since there are many flags in {\TPS}, the flags which are most crucial
to mating search have been collected together under the subject
\indexother{IMPORTANT}, so that typing LIST IMPORTANT will produce a
list of the settings of the dozen or so most important flags for
mating search.

{\TPS} has a facility called \indexcommand{UNIFORM-SEARCH},
which attempts to find the correct flag settings
on its own; it is capable of finding correct settings for many of the theorems
which {\TPS} has proven to date. See section \ref{testtop} for more details.


\section{Review, flags, and modes}


The Review top-level is provided to make it easier to update flag-settings.
Enter the top-level with the command \indexcommand{REVIEW}.  The commands provided
in this top-level (for a listing, enter \indexcommand{?}) give information about the
various flags available.  Each flag is listed under one or more subjects.
You may use the command \indexcommand{SETFLAG} to change the value of a flag, or
you may simply type the name of the flag itself. When entering the argument for
a flag, you may type \indexcommand{??} for more information about the argument
type. In some cases, {\TPS} will warn the user that the flag is irrelevant, given
the settings of other flags.
File names and extensions should be entered as strings (i.e. surrounded
by double quotes); although in some circumstances one can omit the quotes, it
is never an error to include them.

All the flags for a subject can be listed by using the \indexcommand{LIST} command;
{\tt help list} will tell you more. You may search the help messages of flags by
using the \indexcommand{KEY} command; for example {\tt key `unification' all} will
find the flags whose help messages mention unification, in all subjects. A similar
command \indexcommand{SEARCH} allows you to search the help messages of all {\TPS}
objects.

To update many flags at once, use \indexcommand{UPDATE} or 
\indexcommand{UPDATE-RELEVANT}.
The \indexcommand{UPDATE} commmand 
allows the user to set all the flags in given subjects.
Since certain flag values render some other flags
irrelevant, there is a command called \indexcommand{UPDATE-RELEVANT} which
allows users to conveniently take into account the relationships
between flags and the decisions previously made while setting flags
interactively.  Since proof procedures are constantly being developed
or refined, when \indexcommand{UPDATE-RELEVANT} is called, the user is given the
option of using the current flag relevancy information in memory,
loading flag relevancy information which has been saved to a file, or
rebuilding flag relevancy information automatically by processing the Lisp
source files.  

Modes are collections of flags with specified settings.  By selecting a
mode, you simultaneously set all the included flags.  You can also
define your own modes and save them.  See the \indexcommand{INSERT} command,
in the library top level, for more information.

See section \ref{searchanalysis} for information on how to use
a known proof to suggest {\TPS} flag settings, and Chapter
\ref{ms-guide} for information about certain flags which affect the
search processes.

Currently defined flags and modes are listed in the facilities guide
\cite{AndrewsTPS88e}.

\input{testtop}
\input{searchanalysis}
