\chapter{Rewrite Rules and Theories}
\label{rewrite}

{\TPS} allows the user to define rewrite rules, and to apply them in interactive and automatic proofs.

Rewrite rules may be polymorphic and/or bidirectional; there may be a function attached to test for
the applicability of the rule (the default is `always applicable'), and there may also be another function
to be applied after the rule is applied - for example, lambda-normalization
(the default is that there is no such function).

Bidirectional rules are considered to `normally' work left-to-right, but
the user will be prompted if there is any ambiguity. (So, for example,
\indexcommand{APPLY-RRULE} and \indexcommand{UNAPPLY-RRULE} are not
distinguishable for a bidirectional rule, unless it has an associated function.)

Theories are basically collections of rewrite rules and gwffs (and possibly some other previously defined subtheories).
They have %only two uses in {\TPS}.
several uses in {\TPS}.
%Firstly,
They can be saved into the library, so the user can define a theory containing all of
the required axioms and rewrite rules and then load it with a single FETCH command.
%Secondly,
The command \indexcommand{USE-THEORY} activates all of the rewrite rules of the
given theory and deactivates all other rewrite rules in memory, allowing the user to switch
easily between different theories.
One can include the currently active rewrite rules as additional premises into
proofs by setting the flag \indexflag{ASSERT-RRULES} (see
\ref{RRulesInAutoProofs}).
Within the rewriting top level (see Section \ref{RewritingTopLevel}), theories
are used to describe rewrite relations. Theories designed solely for use within
the rewriting top level will typically contain no axioms.
When a theory is loaded from the library, {\TPS} also creates an abbreviation
which is the conjunction of the axioms, the universal
closure of the rewrite rules and the abbreviations representing any subtheories
given by the user. This allows the user to have sentences like
`PA IMPLIES [[ONE PLUS ONE] = TWO]', if PA is a theory.

%The top-level commands for manipulating rewrite rules are as follows:

\section{Top-Level Commands for Manipulating Rewrite Rules}
\begin{description}
\item[] \indexcommand{FETCH} gets an rrule from the library.

\item[] \indexcommand{DELETE-RRULE} deletes an rrule from {\TPS}.

\item[] \indexcommand{LIST-RRULES} lists the rrules currently in memory.

\item[] \indexcommand{MAKE-ABBREV-RRULE} creates an rrule from an abbreviation (eg EQUIVS, above).

\item[] \indexcommand{MAKE-INVERSE-RRULE} creates a new rule which is the inverse of an old rule.

\item[] \indexcommand{MAKE-THEORY} creates a new theory, which can be saved in the library.

\item[] \indexcommand{PERMUTE-RRULES} reorders the list of rrules in {\TPS}. (By default, {\TPS} will try to apply the
active rules in the order in which they are listed.)

\item[] \indexcommand{REWRITE-SUPP1} does one step of rewriting in an ND proof (see APPLY-ANY-RRULE).

\item[] \indexcommand{REWRITE-SUPP*} does many steps in an ND proof (see APPLY-ANY-RRULE*).

\item[] \indexcommand{UNREWRITE-PLAN*} and \indexcommand{UNREWRITE-PLAN1} are the same but in the other direction.
 Because the last four can get confusing, we also have:

\item[] \indexcommand{SIMPLIFY-PLAN}, \indexcommand{SIMPLIFY-PLAN*}, \indexcommand{SIMPLIFY-SUPP} and \indexcommand{SIMPLIFY-SUPP*}
 which attempt to simplify a plan or support line assuming that the left-to-right
  direction is `simplification', and that the higher-numbered lines should
  always be rewrite instances of the lower-numbered lines.
  Now that we have bidirectional rules, this makes sense.
  All of the ND commands use active rules only. All rules are active when
  first loaded/defined, and remain active until deactivated by the user.

\item[] \indexcommand{ACTIVATE-RULES} and \indexcommand{DEACTIVATE-RULES} allow you to turn rules on and off.

\item[] \indexcommand{USE-THEORY} activates all the rewrite rules in the given theory and deactivates
  all other rules. If the theory is unknown, {\TPS} will attempt to load it from the library.

\item[] \indexcommand{USE-RRULES}, and its associated wffop \indexother{INSTANCE-OF-REWRITING}, allow you to deduce
 any wff B(O) from a wff A(O) provided that A rewrites to B *without* using any
 overlapping (nested) rewrite rules. In particular, you can generate one wff from
 the other in the editor, rewriting only those parts you want to rewrite.
\end{description}

%There are also a number of editor operations dealing with rewrite rules:

\section{Editor Operations Dealing with Rewrite Rules}
\begin{description}
\item[] \indexcommand{ARR} is the editor command which applies the first applicable rrule.

\item[] \indexcommand{ARR*} does ARR until it terminates.
  (both of the above use active rules only)

\item[] \indexcommand{ARR1} applies a particular rrule once.

\item[] \indexcommand{ARR1*} applies the same rrule until it terminates.
  (both of the above can use either active or inactive rules).

\item[] \indexcommand{MAKE-RRULE} makes an rrule whose lhs is the current edwff.

\item[] \indexcommand{UNARR}, \indexcommand{UNARR*}, \indexcommand{UNARR1}, \indexcommand{UNARR1*} are the
  editor commands which apply rrules in the reverse direction.
\end{description}
%
%An example of how to use these commands will be given later.

\section{An Example of Rewrite Rules in Interactive Use}

\begin{tpsexample}
<lib25>fetch theo2
TYPE (LIB-ARGTYPE-OR-NIL): argtype or nil [NIL]>
THEO2
{\it We retrieve the theory THEO2 from the library.}
<lib26>help theo2
THEO2 is a theory and a logical abbreviation.
     -----
As a theory:
THEO2 is an extension of (THEO1)
Rewrite rules are: (ONE)
     -----
As a logical abbreviation:

THEO1 \(\and\) \(\one\) = SUCC\(\sb{\greeks\greeks}\) ZERO\(\sb{\greeks}\)
{\it THEO2 contains THEO1 as a subtheory; let's look at that.}
<lib27>help theo1
THEO1 is a theory and a logical abbreviation.
     -----
As a theory:
theory of arithmetic, sort of
Rewrite rules are: (ADD-A ADD-B)
     -----
As a logical abbreviation:
theory of arithmetic, sort of
\(\forall\)x\(\sb{\greeks}\) [x + ZERO\(\sb{\greeks}\) = x] \(\and\) \(\forall\)y\(\sb{\greeks}\) \(\forall\)x.x + SUCC\(\sb{\greeks\greeks}\) y = SUCC.x + y
{\it THEO1 is the usual theory of ZERO, SUCC and +}
<lib28>leave
<29>list-rrules
Currently defined rewrites are: (ONE ADD-B ADD-A)
These are defined as follows:
ONE : \(\one\) <--> SUCC\(\sb{\greeks\greeks}\) ZERO\(\sb{\greeks}\)
ADD-B : x\(\sb{\greeks}\) + SUCC\(\sb{\greeks\greeks}\) y\(\sb{\greeks}\) <--> SUCC\(\sb{\greeks\greeks}\).x\(\sb{\greeks}\) + y\(\sb{\greeks}\)
ADD-A : x\(\sb{\greeks}\) + ZERO\(\sb{\greeks}\) <--> x\(\sb{\greeks}\)

Of these, (ONE ADD-B ADD-A) are active.
{\it We see that there are now three rewrite rules, all active and all bidirectional.}
<30>ed sum1
<Ed31>p
SUCC\(\sb{\greeks\greeks}\) ZERO\(\sb{\greeks}\) + ZERO +.SUCC [SUCC.SUCC ZERO] + SUCC [SUCC.SUCC ZERO] + ZERO
{\it Now let's try to reduce this expression.}
<Ed32>arr*
Apply bidirectional rules in the forward direction? [Yes]>
SUCC\(\sb{\greeks\greeks}\).SUCC.SUCC.SUCC.SUCC.SUCC.SUCC ZERO\(\sb{\greeks}\)
{\it That was perhaps a little quick; let's do it again more slowly.}
<Ed33>sub sum1
SUCC\(\sb{\greeks\greeks}\) ZERO\(\sb{\greeks}\) + ZERO +.SUCC [SUCC.SUCC ZERO] + SUCC [SUCC.SUCC ZERO] + ZERO
<Ed34>arr
Apply bidirectional rules in the forward direction? [Yes]>
SUCC\(\sb{\greeks\greeks}\) ZERO\(\sb{\greeks}\) + ZERO +.SUCC [SUCC [SUCC.SUCC ZERO] + SUCC.SUCC ZERO] + ZERO
{\it Here we took the first applicable rewrite rule.}
<Ed35>arr1*
RULE (SYMBOL): name of rule [No Default]>add-b
Apply rule in the forward direction? [Yes]>
SUCC\(\sb{\greeks\greeks}\) ZERO\(\sb{\greeks}\) + ZERO +.SUCC [SUCC.SUCC.SUCC [SUCC.SUCC ZERO] + ZERO] + ZERO
{\it Then we apply ADD-B as much as possible.}
<Ed36>arr1*
RULE (SYMBOL): name of rule [No Default]>add-a
Apply rule in the forward direction? [Yes]>
SUCC\(\sb{\greeks\greeks}\) ZERO\(\sb{\greeks}\) + SUCC.SUCC.SUCC.SUCC.SUCC.SUCC ZERO
{\it Then ADD-A...}
<Ed37>arr1*
RULE (SYMBOL): name of rule [No Default]>add-b
Apply rule in the forward direction? [Yes]>
SUCC\(\sb{\greeks\greeks}\).SUCC.SUCC.SUCC.SUCC.SUCC.SUCC ZERO\(\sb{\greeks}\) + ZERO
{\it Then ADD-B, and clearly one more ADD-A would do it}
<Ed38>arr1*
RULE (SYMBOL): name of rule [No Default]>ONE
Apply rule in the forward direction? [Yes]>no
SUCC\(\sb{\greeks\greeks}\).SUCC.SUCC.SUCC.SUCC.SUCC.\(\one\) + ZERO\(\sb{\greeks}\)
{\it ...but instead we opt to apply ONE from right to left.}
<Ed39>ok
<40>prove sum3
SUM3 is not a known gwff. Search for it in library? [Yes]>
PREFIX (SYMBOL): Name of the Proof [No Default]>sum3
NUM (LINE): Line Number for Theorem [100]>
(100)  \(\assert\)   \(\one\) + \(\one\) + [ZERO\(\sb{\greeks}\) + SUCC\(\sb{\greeks\greeks}\) \(\one\)] + SUCC [ZERO + \(\one\)]
          = SUCC .SUCC .SUCC .SUCC .SUCC .SUCC ZERO & PLAN1
{\it Now let's try to prove a theorem with these rewrite rules.}
<41>simplify-plan*
P2 (LINE): Line after rewriting (higher-numbered) [100]>
P1 (LINE): Line before rewriting (lower-numbered) [99]>
(99)   \(\assert\)   SUCC\(\sb{\greeks\greeks}\) [SUCC .SUCC .SUCC .SUCC .SUCC ZERO\(\sb{\greeks}\)]
          = SUCC .SUCC .SUCC .SUCC .SUCC .SUCC ZERO & PLAN2
{\it All we need do now is ASSERT REFL=. Instead, we'll delete this line and prove it more slowly.}
<42>delete 99
<43>simplify-plan
P2 (LINE): Line after rewriting (higher-numbered) [100]>
P1 (LINE): Line before rewriting (lower-numbered) [99]>
(99)   \(\assert\)   SUCC\(\sb{\greeks\greeks}\) ZERO\(\sb{\greeks}\) + \(\one\) + [ZERO + SUCC \(\one\)] + SUCC [ZERO + \(\one\)]
          = SUCC .SUCC .SUCC .SUCC .SUCC .SUCC ZERO & PLAN4
<44>simplify-plan
P2 (LINE): Line after rewriting (higher-numbered) [99]>
P1 (LINE): Line before rewriting (lower-numbered) [98]>
(98)   \(\assert\)   SUCC\(\sb{\greeks\greeks}\) ZERO\(\sb{\greeks}\) + SUCC ZERO + [ZERO + SUCC \(\one\)] + SUCC [ZERO + \(\one\)]
          = SUCC .SUCC .SUCC .SUCC .SUCC .SUCC ZERO & PLAN5
{\it We could continue in this way, or we can try restricting the range of rewrite rules to be applied, as follows:}
<45>deactivate-rules
RLIST (RRULELIST): Rules to deactivate [(ONE ADD-B ADD-A)]>ADD-B ADD-A
{\it The only rule left to apply is now ONE.}
<46>simplify-plan*
P2 (LINE): Line after rewriting (higher-numbered) [98]>
P1 (LINE): Line before rewriting (lower-numbered) [97]>
(97)   \(\assert\)     SUCC\(\sb{\greeks\greeks}\) ZERO\(\sb{\greeks}\) + SUCC ZERO + [ZERO + SUCC .SUCC ZERO]
            + SUCC [ZERO + SUCC ZERO]
          = SUCC .SUCC .SUCC .SUCC .SUCC .SUCC ZERO & PLAN6
{\it Every instance if ONE is now rewritten to SUCC ZERO}
<47>activate-rules
RLIST (RRULELIST): Rules to activate [(ONE ADD-B ADD-A)]>ADD-B
{\it We reactivate ADD-B, which is the rule taking (x+(SUCC y)) to SUCC(x+y).}
<48>simplify-plan*
P2 (LINE): Line after rewriting (higher-numbered) [97]>
P1 (LINE): Line before rewriting (lower-numbered) [96]>
(96)   \(\assert\)   SUCC\(\sb{\greeks\greeks}\) [SUCC .  SUCC [SUCC .SUCC [SUCC ZERO\(\sb{\greeks}\) + ZERO] + .ZERO + ZERO]
                          + .ZERO + ZERO]
          = SUCC .SUCC .SUCC .SUCC .SUCC .SUCC ZERO & PLAN7
<49>activate-rules
RLIST (RRULELIST): Rules to activate [(ONE ADD-B ADD-A)]>
{\it Finally, we reactivate ADD-A, which takes (x+ZERO) to x.}
<50>simplify-plan*
P2 (LINE): Line after rewriting (higher-numbered) [96]>
P1 (LINE): Line before rewriting (lower-numbered) [95]>
(95)   \(\assert\)   SUCC\(\sb{\greeks\greeks}\) [SUCC .SUCC .SUCC .SUCC .SUCC ZERO\(\sb{\greeks}\)]
          = SUCC .SUCC .SUCC .SUCC .SUCC .SUCC ZERO & PLAN8
<51>assert refl=
LINE (LINE): Line with Theorem Instance [No Default]>95
{\it ...and the proof is done.}
\end{tpsexample}

\section{Using Rewrite Rules in Automatic Proof Search}
\label{RRulesInAutoProofs}

The \indexother{MS98-1} search procedure can extract rewrite rules from wffs
and use them for proof search. To use this feature within \index{MS98-1}, set
the flag \indexflag{MS98-REWRITES}. You may also need to set the flag
\indexflag{MAX-SUBSTS-VAR} to some positive value.

By default the procedure will not use any rewrite rules apart from those it
can extract during proof search. In some cases, though, it may be beneficial
to allow the procedure using additional rewrite rules. This can be done in two
ways.

The recommended way to add rewrite rules is adding them as equational premises
to the main assertion. To add the active rewrite rules as premises to the
assertion of a proof, set the flag \indexflag{ASSERT-RRULES} before beginning
a new proof using \indexcommand{PROVE}. Assume, for instance, you have two
active rewrite rules (of the form \mbox{\texttt{$l_i$ <--> $r_i$}} for
$i\in\{1,2\}$). After setting \indexflag{ASSERT-RRULES}, you start a proof by
entering \texttt{PROVE "\textit{main-assertion}"}. The resulting assertion will
be of the form \texttt{$l_2$ = $r_2$ $\implies$ .$l_i$ = $r_i$ $\implies$ \textit{main-assertion}}.

\indexother{MS98-1} may fail to recognize some of the rules passed by the above
method. This usually indicates that the rule is too complex to be used by the
procedure.
If you want to enforce that all active rewrite rules are indeed used by the
search procedure, you can set the flag \indexflag{MS98-EXTERNAL-REWRITES}.
This flag works independently from the setting of \indexflag{ASSERT-RRULES}.
Normally, i.e. when the flag is not set, \indexother{MS98-1} temporarily
deactivates all active rewrite rules which were not extracted from the
assertion to prove. When \indexflag{MS98-EXTERNAL-REWRITES} is set, the
globally active rewrite rules remain active and are used in addition to the
rules extracted from the proof assertion.

\input{rewrite-top}