\chapter{ETPS}\label{etps}

\section{The Outline Modules}

The {\tt OUTLINE} modules in TPS have two main subparts; 
the bookkeeping functions, and the \indexcommand{GO} command which
gives sophisticated help or constructs a proof automatically.
They are collected in the modules of the form {\tt OTL*}.

In ETPS only the bookkeeping functions are present.

The discussion below is aimed at understanding the {\tt OUTLINE} modules
independently of the system, but we generally assume we are working in
\ETPS.  If \TPS differs, this is noted.

We often talk about {\it proofs}, even though they are properly only incomplete
proofs or {\it proof outlines}.  It is assumed that the reader knows what
planned lines ({\it plines}) and deduced lines ({\it dlines}) are.  This and 
general familiarity with \ETPS are necessary to understand this discussion.

\subsection{Proofs as a Data Structure}

Proofs in \ETPS are represented by a single atom with a variety of properties.
The global variable {\tt DPROOF} has as value the name of the current proof.
In case you are working, say on exercise {\tt X6200}, {\tt DPROOF} will have the 
value {\tt X6200}.  The current proof name then has a variety of properties.

\begin{description}
\item [{\tt LINES}] {\tt ({\it line} {\it line} ...)}.  This is simply an ordered list of all
lines in the current proof without repetition.  The order is such that
lines with a lower number appear first in the list.
\\ {\bf WARNING}:  This property is frequently changed destructively.
As a consequence it may never be empty and should be used for other
purposes only in a copy.

\item [{\tt LINEALIASES}] {\tt (({\it line} . {\it no})({\it line} . {\it no}) ...)}.  This is an 
unordered association list correlating lines with their numbers.  No line 
should ever appear in more than one pair, and neither should a number.  Try to 
think of arguments for and against this representation, compared to one where
the number of a line is stored on the line's property list.
\\ {\bf WARNING}:  This property is frequently changed destructively.
As a consequence it may never be empty and should be used for other
purposes only in a copy.

\item [{\tt PLANS}] {\tt {(({\it pline} . {\it supportlist})({\it pline} . {\it supportlist}) ...)}}.
This stores the important {\it plan}-{\it support} {\it structure} of the current
proof.  {\it pline} is a still unjustified line in the current proof,
{\it supportlist} is a list of deduced lines supporting a {\it pline}.
A {\it pline} may never have a justification other than {\tt PLAN}{\it i}, 
a {\it sline} (support line) must always be completely justified, i.e.
may not ultimately depend on a planned line.

The association list is ordered such that the most recently affected
{\it pline} is closer to the front of the list.  The order can be changed
explicitly with the {\tt SUBPROOF} command.

{\bf WARNING}:  This property is frequently changed destructively.
As a consequence it may never be empty and should be used for other
purposes only in a copy.

\item [{\tt GAPS}] ({\it gap} {\it gap} ...). This is a list of the gaps between lines in the
proof. Each gap has the properties ({\tt MIN-LABEL} {\it line}) and ({\tt MAX-LABEL} {\it line}).

\item [{\tt NEXTPLAN-NO}] {\it integer}. This is just the next number that will be used for 
a planned line.

\item [{\tt ASSERTION}] {\it gwff}. This is the assertion being proven.
\end{description}

When a proof is saved using the \indexcommand{SAVEPROOF} command, a checksum may be generated.
This is used by \ETPS to verify that the saved proof has not been manually edited by a student
(otherwise it would be possible to edit out the planned lines and convince \ETPS to issue the 
\indexcommand{DONE} command). Since it takes time to generate the checksum, it is only 
generated if the flag \indexflag{EXPERTFLAG} is NIL. This means that proofs written by \TPS
with \indexflag{EXPERTFLAG} T cannot be read into \ETPS with \indexflag{EXPERTFLAG} NIL.


\subsection{Proof Lines as a Data Structure}

Proof lines in \ETPS have a variety of properties:

\begin{description}
\item [{\tt REPRESENTS}] This is the wff asserted by the line.  In the original 
TPS this had to be an
atomized wff of very particular structure, which lead to numerous problems
in higher-order logic.  In \ETPS this has been maintained for the present. Our
goal, of course, is to allow arbitrary {\it gwffs} as {\tt REPRESENTS} of lines.

\item [{\tt HYPOTHESES}] This is a list of lines assumed as hypotheses for the line.
The list of hypotheses is ordered (lowest numbered line first), but to
my knowledge no function assumes this.  It simply looks better in the output.
No line should appear twice as an hypothesis (this fact may actually be used
here and there).

\item [{\tt JUSTIFICATION}]  {\tt {{\it RULE} {\it gwfflist} {\it linelist}}} The line can be
inferred by an inference rule {\it RULE}from {\it linelist}.  
{\it gwfflist} has somehow been used to infer the line.

\item [{\tt LINENUMBER}] The line number associated with the line.
\end{description}

\input{outline}
